Modular Model-Checking of a Byzantine Fault-Tolerant Protocol

### 2016-12-08 doc/om-paper/paper.tex
edits throughout.

\title{Modular Model-Checking Implementations of a Byzantine Fault-Tolerant Protocol}

@@

\begin{abstract}
With proof techniques like IC3 and $k$-induction, model-checking scales further
than ever before.  Still, fault-tolerant distributed systems are particularly
challenging to model-check given their large state space and
non-determinism. The typical approach to controlling complexity is to construct
ad-hoc abstractions of faults, message-passing, and behaviors.  However, these
abstractions come at the price of divorcing the model from its implementation
and making refactoring difficult. In this work, we present a framework for
fault-tolerant distributed system verification that combining ideas from the literature including calendar automata,
symbolic fault injection, and abstract state machines observers, and then we apply the framework to model-checking various implementations of the Hybrid Oral Messages algorithm that differ in the fault model, timing model, and local node behavior. We show that despite being low-level models, the verifications are scalable and modular, insofar as isolated changes to an implementation require modular changes to the model and proofs.
This work is carried out in SAL model-checker.
\end{abstract}

\lee{grep for synchronous observers section references}

@@

In this paper, we present a framework for modeling fault-tolerant distributed systems, and use that framework to verify several implementations of the Byzantine fault-tolerant Hybrid Oral Messages algorithm (\OMH)~\cite{Lincoln-Rushby}. The framework combines ideas from the literature and folklore to build scalable and modular formal models suitable for infinite-state model-checking, and it reduces the need for ad-hoc abstractions and optimizations while maintaining scalability. In Section~\ref{sec:model}, we present the important aspects of the framework, including \emph{calendar automata}, originally developed by Dutertre and Sorea~\cite{cal}, symbolic fault-injection, and abstract state machines for verification~\cite{}.

@@

We use the framework to verify low-level models of \OMH in which message passing is explicit, nodes are not enforced to execute strictly synchronously, and voting is explicit. In short, the models corresponds closely with an implementation of the algorithm. In Section~\ref{sec:byz}, we describe \OMH, an initial implementation model of it using the Boyer-Moore Fast Majority Vote algorithm~\cite{mjrty}, and a class of modular invariants; that is, invariants are grouped in a small set of classes with well-defined relationships. The verification is interesting in its own right, as it is the first fully parametric (on the number of nodes) model-checked \emph{implementation} of the algorithm.

In Section~\ref{sec:experimental}, we first show that despite being low-level, the model is scalable. Developing invariants requires some user guidance, and isolated changes to an implementation should require isolated modifications to the model and proof. To demonstrate this, we modify the \OMH implementation along the dimensions of faults (by adding an omissive-asymetric fault type~\cite{omissive}), time (by making a time-triggered model), and local behavior (by changing the majority vote to a mid-value selection) and show that in each case, the modifications are small and modular.

@@

Proofs of these two lemmas are straightforward so we omit them here.

@@

\subsection{Symbolic Fault Injection: a Synchronous Kibitzer}\label{sec:kibitzer}
The typical approach to modeling faults is to add new state variables to each process representing its fault state. Then additional state variables are added to represent the non-determinism introduced by faults into the state or messages by a process. As a simple example, we might define a node that sends a good message if it is non-faulty and a bad message otherwise. In pseudo-code using guarded commands, its definition might look like the following:

@@

\noindent
But this typical approach mixes the specification of a node's behavior with the fault model, an aspect of the environment. Generally, nodes do not contain state variables assigned to their faults, or use their fault-status to determine their behavior!\footnote{There are exceptions. Benign faults may be detected by a node itself (e.g., in a built-in-test).} The upshot is that combining faults and node state divorces the specification from its implementation.

@@

In order to scale verification of models up to more than a few nodes, and to avoid having to increase the depth of induction (which tends to be very expensive) during verification, we have followed Rushby's  application of abstract transition systems (and disjunctive invariants) \cite{Rushby00:CAV,timed-systems}. In this context, an abstract transition system, relative to a given transition system $\mathcal{M} = (S, I, \rightarrow)$, is a set of state predicates $A_1, \ldots, A_n$ over $S$ and a transition system $\tld{\mathcal{M}} = (\tld{S}, \tld{I}, \leadsto)$ such that:

@@

The Hybrid Oral Messages ($\OMH$) algorithm~\cite{Lincoln-Rushby} is a variant of the classic Oral Messages ($\OM$) algorithm~\cite{om}, originally developed by Thambidurai and Park~\cite{hybrid} to achieve distributed consensus in the presence of a hybrid fault model. However, $\OMH$ had a bug, as originally formulated, which was corrected and the mended algorithm was formally verified by Lincoln and Rushby using interactive theorem-proving~\cite{Lincoln-Rushby}.

@@

We described a hybrid fault model in Section~\ref{sec:kibitzer}. Under that fault model, agreement and validity hold if $2a+2s+b+1 \geq n$, where $n$ is the total number of nodes, $a$ is the number of Byzantine (or asymmetric) faults, $s$ is the number of symmetric faults, and $b$ is the number of benign faults. Additionally, it is required that the number of rounds $m$ be greater or equal to the number of Byzantine faults, $a$~\cite{Lincoln-Rushby,hybrid}.

@@

We have implemented $\OMH(1)$ (as well as the variants described in Section~\ref{sec:modular}) in Symbolic Analysis Laboratory (SAL)~\cite{SAL}. SAL contains a suite of model-checkers. In our work, we use infinite-state (SMT-based) $k$-induction~\cite{cal}.

@@

To make the proof scalable, we specify inductive invariants to be used by SAL's $k$-induction engine. There are 11 invariants, falling into five categories:
\begin{enumerate}
  \item \emph{Calendar automata}: Lemmas relating to the calendar automata model. These include lemmas such as time being monotonic, channels missing messages if there is no calendar event, and only nodes associated with a calendar event may execute their local state machines.
  \item \emph{Abstract state machine (ASM)}: Lemmas relating the ASM synchronous observer states to the implementation states.
  \item \emph{Lieutenant local state-machine (SM) behavior}: Lemmas describing the modes of behavior of the lieutenants. The major modes of the state machine are receiving messages, then once it has filled its buffer, it votes, and after voting returns the result. One additional lemma notes that the messages currently received plus missing messages equals the total number of expected messages.
  \item \emph{Faults}: Lemmas characterizing the effect of a fault in a single broadcast. Examples include lemmas stating that if a node receives a faulty message, some ``upstream'' node in the communication path was faulty. Another example is that the faults of messages latched by a node in its buffer match the faults ascribed to the sender in the calendar event.
  \item \emph{Voting}: Lemmas proving that the Fast Majority Vote algorithm implements a majority vote, if one exists. These lemmas are nearly verbatim transcriptions from the journal proofs~\cite{mjrty}.
\end{enumerate}
\noindent
The proof structure is shown in Figure~\ref{fig:proof}. The number of lemmas per category are shown in parentheses. Arrows denote dependencies. For example, the ASM lemmas depend on both the calendar automata and lieutenant state-machine lemmas. As can be seen, the proof structure is modular. The calendar lemmas are general and independent of any particular protocol or fault model. Similarly, lemmas about the internal behavior of a lieutenant is independent of the global protocol behavior. It is also independent of the effect of faults on the system---the only ``knowledge'' of faults that lieutenant has is whether a fault is benign or not. Lemmas about the behavior of faults in the system are also independent of the particular protocol being modeled. Likewise, lemmas about the particular voting algorithm used depend only on the lieutenant's internal behavior. Only the ASM depends on both calendar-specific and local state-machine results, since it is an abstraction of the entire system implementation. Recall, however, that the ASM is a convenience for debugging and can be elided.

@@

As a point of comparison, Rushby presents an elegant model of the simpler \OM algorithm, also in SAL, but using BDD-based model-checking~\cite{Rushby:OM1}. The model is at the algorithmic level. For small values, the verification is much faster, probably due to making only one call to SAL. However, for six relays and two receivers, it takes 449~seconds and timeouts for seven relays and two receivers.

@@

To demonstrate the modularity of the model and verification approach, in this section, we explore variants to the model and report the effort required to implement the modifications and repair the proofs. The results are summarized in Figure~\ref{fig:effort} and sketched below. In the table, for each modification, we report how much of the model must be modified. We report on four aspects of the system: which transition systems making up the full model are modified (as described in Section~\ref{sec:model}), how many definitions have to be added or modified, the number of invariants that have to be added or modified, and which invariant classes (as defined in Section~\ref{sec:invariants}) those lemmas belong to. We modify the implementation along the axes of faults, time, and local node behavior.

@@

Adding new kinds of faults requires more work but is still modular. Consider adding \emph{omissive asymmetric faults}, a restriction of Byzantine faults in which a broadcaster either sends the correct value or a benign fault~\cite{omissive}, to the fault model. Doing so requires modifying none of transition systems. We add a new uninterpreted function definition for omissive asymmetric faults, then modify the type of faults, and their effect on the calendar. Two invariants, both in the class of invariants cover faults, are extended to cover the cases where a sender is omissive asymmetric.

\paragraph{Time-Triggered Messaging.}
A \emph{time-triggered} distributed system is one in which nodes are independently clocked, but clock constraints (ensured by periodic clock synchronization) allow the model to appear as if it is executing synchronously~\cite{kopetz}.

Changing the model to be time-triggered principally requires making the source, relays, and receivers driven explicitly by the passage of time. As well, a ``receive window'' is defined at which messages from non-faulty nodes should be received. Messages received outside the window are marked as coming from manifest-faulty senders. To implement a time-triggered model, we create three definitions: encoding nondeterministic message delay, and two definitions that are helpers in invariants. The guards in the relays and receivers are modified to latch messages received outside the specified windows as being manifest faults. Additionally, the ASM definition is modified to track the times in the calendar, not just the messages. Two new invariants calendar invariants are introduced, stating that the calendar messages are either empty, or their time-stamps fall within the respective message windows. Then, three invariants classifying faults are relaxed to allow for the possibility of faulty nodes sending benign messages.

\paragraph{Mid-Value Selection.}
Our $\OMH(1)$ model leverages majority vote in order to tolerate faults. Another choice for the fault masking algorithm used here is mid-value selection (or median). This choice is more common in applications involving hardware, signal selection (where signal values cannot be relied on to exactly match even in the absence of faults), or cases where information about congruence is useful. To implement mid-value selection in our model, we allow messages sent to take values in $\mathbb{R}$ and the receiver transition system is modified in two ways. First, a second buffer is introduced which will hold the sorted contents of the main buffer once voting has commenced. Second, a mid-value select function is called on the sorted buffer and the result is stored as the receiver's vote. The only invariants needing modification were the ASM definition (to account for the values stored by the new buffer and the relation between it and the main buffer) and the voting invariant.

@@

As shown in Section~\ref{sec:modular}, modifications to the implementation are modular. In particular, any protocol specified in the model can reuse the lemmas associated with the calendar automata and fault-model (insofar as they use the same fault model).

>>>>

\title{Modular Model-Checking of a Byzantine Fault-Tolerant Protocol}

@@

\begin{abstract}
With proof techniques like IC3 and $k$-induction, model-checking scales further
than ever before.  Still, fault-tolerant distributed systems are particularly
challenging to model-check given their large state spaces and
non-determinism. The typical approach to controlling complexity is to construct
ad-hoc abstractions of faults, message-passing, and behaviors.  However, these
abstractions come at the price of divorcing the model from its implementation
and making refactoring difficult. In this work, we present a framework for
fault-tolerant distributed system verification that combining ideas from the literature including calendar automata,
symbolic fault injection, and abstract state machines observers, and then we apply the framework to model-checking various implementations of the Hybrid Oral Messages algorithm that differ in the fault model, timing model, and local node behavior. We show that despite being implementation-level models, the verifications are scalable and modular, insofar as isolated changes to an implementation require modular changes to the model and proofs.
This work is carried out in SAL model-checker.
\end{abstract}

\lee{grep for synchronous observers section references}
\lee{grep for ASM vs abstract transition system}

@@

In this paper, we present a framework for modeling fault-tolerant distributed systems, and use that framework to verify several variant implementations of the Byzantine fault-tolerant Hybrid Oral Messages algorithm (\OMH)~\cite{Lincoln-Rushby}. The framework combines ideas from the literature and folklore to build scalable and modular formal models suitable for infinite-state model-checking, and it reduces the need for ad-hoc abstractions and optimizations while maintaining scalability. In Section~\ref{sec:model}, we present the important aspects of the framework, including calendar automata, originally developed by Dutertre and Sorea~\cite{cal}, symbolic fault-injection, and abstract state machines for verification.

@@

We use the framework to verify implementation-level models of \OMH in which message passing is explicit, nodes are not enforced to execute strictly synchronously, and voting is explicit. In short, the models corresponds closely with an implementation of the algorithm. In Section~\ref{sec:byz}, we first describe \OMH, then an implementation of it that uses the Boyer-Moore Fast Majority Vote algorithm~\cite{mjrty}. We then describe a set of modular invariants, such that the invariants only concern specific aspects of the model (e.g., faults, local node behavior, or the passage of time). The verification is interesting in its own right, as it is the first fully parametric (on the number of nodes) model-checked \emph{implementation} of the algorithm.

In Section~\ref{sec:experimental}, we first show that despite being implementation-level, the model is scalable. Developing invariants requires some user guidance, and isolated changes to an implementation should require isolated modifications to the model and proof. To demonstrate this, we modify the \OMH implementation along the dimensions of faults (by adding an omissive-asymetric fault type~\cite{omissive}), time (by making a time-triggered model), and local behavior (by changing the majority vote to a mid-value selection) and show that in each case, the modifications are small and modular.

@@

Proofs of these two lemmas are straightforward and omitted.

@@

\subsection{Symbolic Fault Injection: a Synchronous Kibitzer}\label{sec:kibitzer}
The typical approach to modeling faults is to add new state variables to each process representing its fault state. Then a node chooses actions based on its fault state. As a simple example, we might define a node that sends a good message if it is non-faulty and a bad message otherwise. In pseudo-code using guarded commands, its definition might look like the following:

@@

\noindent
But this approach mixes the specification of a node's behavior with the fault model, an aspect of the environment. Generally, nodes do not contain state variables assigned to their faults, or use their fault-status to determine their behavior!\footnote{There are exceptions; for example, benign faults may be detected by a node itself (e.g., in a built-in-test).} The upshot is that combining faults and node state divorces the specification from its implementation.

@@

Due to the sheer size of implementation-level models, manually examining counterexamples is tedious. To scale up verification, we use abstract transition systems (also known as disjunctive invariants)~\cite{Rushby00:CAV,timed-systems}. In this context, an abstract transition system, relative to a given transition system $\mathcal{M} = (S, I, \rightarrow)$, is a set of state predicates $A_1, \ldots, A_n$ over $S$ and a transition system $\tld{\mathcal{M}} = (\tld{S}, \tld{I}, \leadsto)$ such that:

@@

The Hybrid Oral Messages ($\OMH$) algorithm~\cite{Lincoln-Rushby} is a variant of the classic Oral Messages ($\OM{m}$) algorithm~\cite{om}, originally developed by Thambidurai and Park~\cite{hybrid} to achieve distributed consensus in the presence of a hybrid fault model. However, $\OMH$ had a bug, as originally formulated, which was corrected and the mended algorithm was formally verified by Lincoln and Rushby using interactive theorem-proving~\cite{Lincoln-Rushby}.

@@

We described a hybrid fault model in Section~\ref{sec:kibitzer}. Under that fault model, agreement and validity hold if $2a+2s+b+1 \geq n$, where $n$ is the total number of nodes, $a$ is the number of Byzantine (or asymmetric) faults, $s$ is the number of symmetric faults, and $b$ is the number of benign faults. Additionally, the number of rounds $m$ must be greater or equal to the number of Byzantine faults, $a$~\cite{Lincoln-Rushby,hybrid}.

@@

We have implemented $\OMH(1)$ (as well as the variants described in Section~\ref{sec:modular}) in the Symbolic Analysis Laboratory (SAL)~\cite{SAL}. SAL contains a suite of model-checkers. In our work, we use infinite-state (SMT-based) $k$-induction~\cite{cal}.

@@

To make the proof scalable, we specify inductive invariants to be used by SAL's $k$-induction engine. There are 11 invariants, falling into five categories:
\begin{enumerate}
  \item \emph{Calendar automata}: Lemmas relating to the calendar automata model. These include lemmas such as time being monotonic, channels missing messages if there is no calendar event, and only nodes associated with a calendar event may execute their local state machines.
  \item \emph{Abstract state machine (ASM)}: Lemmas relating the ASM synchronous observer states to the implementation states.
  \item \emph{Receiver local behavior}: Lemmas describing the modes of behavior of the receivers. The major modes of the state machine are receiving messages, then once it has filled its buffer, it votes, and after voting returns the result. One additional lemma notes that the messages currently received plus missing messages equals the total number of expected messages.
  \item \emph{Faults}: Lemmas characterizing the effect of a fault in a single broadcast. Examples include lemmas stating that if a node receives a faulty message, some ``upstream'' node in the communication path was faulty. Another example is that the faults of messages latched by a node in its buffer match the faults ascribed to the sender in the calendar event.
  \item \emph{Voting}: Lemmas proving that the Fast Majority Vote algorithm implements a majority vote, if one exists. These lemmas are nearly verbatim transcriptions from the journal proofs~\cite{mjrty}.
\end{enumerate}
\noindent
The proof structure is shown in Figure~\ref{fig:proof}. The number of lemmas per category are shown in parentheses. Arrows denote dependencies. For example, the ASM lemmas depend on both the calendar automata and receiver state-machine lemmas. As can be seen, the proof structure is modular. The calendar lemmas are general and independent of any particular protocol or fault model. Similarly, lemmas about the internal behavior of a receiver is independent of the global protocol behavior. It is also independent of the effect of faults on the system---the only ``knowledge'' of faults that receiver has is whether a fault is benign or not. Lemmas about the behavior of faults in the system are also independent of the particular protocol being modeled. Likewise, lemmas about the particular voting algorithm used depend only on the receiver's internal behavior. Only the ASM depends on both calendar-specific and local state-machine results, since it is an abstraction of the entire system implementation. Recall, however, that the ASM is a convenience for debugging and can be elided.

@@

As a point of comparison, Rushby presents an elegant model of the simpler \OM{m} algorithm, also in SAL, but using BDD-based model-checking~\cite{Rushby:OM1}. The model is at the algorithmic level. For small values, the verification is much faster, probably due to making only one call to SAL. However, for six relays and two receivers, it takes 449~seconds and timeouts (at one hour) for seven relays and two receivers.

@@

To demonstrate the modularity of the model and verification approach, in this section, we explore variants to the model and report the effort required to implement the modifications and repair the proofs. The results are summarized in Figure~\ref{fig:effort} and sketched below. In the table, for each modification, we report how much of the model must be modified. We report on four aspects of the system: which transition systems are modified (as described in Section~\ref{sec:sketch}), how many definitions have to be added or modified, the number of invariants that have to be added or modified, and which invariant classes (as defined in Section~\ref{sec:invariants}) those lemmas belong to. We modify the implementation along the axes of faults, time, and local node behavior.

@@

Adding new kinds of faults requires more work but is still modular. Consider adding \emph{omissive asymmetric faults}, a restriction of Byzantine faults in which a broadcaster either sends the correct value or a benign fault~\cite{omissive}, to the fault model. Doing so requires modifying none of transition systems, because of the synchronous kibitzer. We add a new uninterpreted function definition for omissive asymmetric faults, then modify the type of faults, and their effect on the calendar. Two invariants, both in the class of invariants cover faults, are extended to cover the cases where a sender is omissive asymmetric.

\paragraph{Time-Triggered Messaging.}
A \emph{time-triggered} distributed system is one in which nodes are independently clocked, but clock constraints allow the model to appear as if it is executing synchronously~\cite{kopetz}.

Changing the model to be time-triggered principally requires making the source, relays, and receivers driven explicitly by the passage of time. As well, a ``receive window'' is defined at which messages from non-faulty nodes should be received. Messages received outside the window are marked as coming from manifest-faulty senders. To implement a time-triggered model, we create three definitions; one encodes nondeterministic message delay, and two are small helpers called by invariants. The guards in the relays and receivers are modified to latch messages received outside the specified windows as being manifest faults. Additionally, the ASM definition is modified to track the times in the calendar, not just the messages. Two new calendar invariants are introduced, stating that the calendar messages are either empty, or their time-stamps fall within the respective message windows. Then, three invariants classifying faults are relaxed to allow for the possibility of faulty nodes sending benign messages.

\paragraph{Mid-Value Selection.}
Our $\OMH(1)$ model leverages a majority vote in order to tolerate faults. Another choice for the fault masking algorithm used is mid-value selection. This choice is common in applications involving hardware, signal selection, or cases where information about congruence is useful. To implement mid-value selection in our model, we allow messages sent to take values in $\mathbb{R}$ and the receiver transition system is modified in two ways. First, a second buffer is introduced which will hold the sorted contents of the main buffer once voting has commenced. Second, a mid-value select function is called on the sorted buffer and the result is stored as the receiver's vote. The only invariants needing modification were the ASM definition (to account for the values stored by the new buffer and the relation between it and the main buffer) and the voting invariant.

@@

The invariants and model are surprisingly modular. The modifications we explored took at most hours to develop. Moreover, most of the invariants do not concern the specific protocol modeled at all, and we hypothesize that for completely different protocols, only the modeling aspects related to the protocol behavior and local node behavior would change.
